---
title: "Code Component"
output:
  html_document: default
  word_document: default
date: "16th of May, 2025"
author: "Laura Larens"
---

# Data Preprocessing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load necessary libraries
library(tidyverse) #(Wickham et al., 2019)
library(readxl) #(Wickham & Bryan, 2025)
library(lme4) #(Bates et al., 2015)
library(ggplot2) #(Wickham, 2016)
```
## Linguistic Data

### Getting Zipf from SUBTLEX-NL (Brysbaert,2020)
```{r}
library(dplyr) #(Wickham et al., 2023)

word_info <- read_excel("experiment_strings.xlsx")

subtlex <- read_excel("SUBTLEX-NL with pos and Zipf.xlsx") %>%
  select(Word, Zipf) %>%
  rename(
    spelling      = Word,
    subtlex_zipf = Zipf
  )

# Joining, adding Zipf to strings, words with no zipf and pseudowords get 1.3555
word_info <- word_info %>%
  left_join(subtlex, by = "spelling") %>%
  mutate(
    Zipf = coalesce(subtlex_zipf, 1.3555)
  ) %>%
  select(-subtlex_zipf)

print(head(word_info))


```




```{r}
# Make is_word into binary values, 1 for word and 0 for pseudoword
word_info <- word_info %>%mutate(is_word_binary = if_else(is_word == "W", 1, 0))

# Add is_complex binary column, 1 morpheme = not complex (0), 2 or more = complex (1)
word_info <- word_info %>% mutate(is_complex = if_else(morph > 1, 1, 0))
```

## Participants' Data

```{r}
#Each participant has a csv file with their data
files <- list.files(pattern = "\\.csv$", full.names = TRUE)


all_data <- map_df(seq_along(files), function(i) {
  file <- files[i]
  file_id <- paste0("P", str_pad(i, 2, pad = "0"))  # Makes sure each file has unique ID



  df <- read_csv(file, show_col_types = FALSE, name_repair = "unique_quiet")

  df <- df %>% filter(!is.na(rt)) #if there is no RT, filters the row out 

  #if pressed_early or no_press is TRUE, filters the row out
  if ("pressed_early" %in% names(df)) {
    df <- df %>% filter(is.na(pressed_early) | pressed_early == FALSE)
  }
  if ("no_press" %in% names(df)) {
    df <- df %>% filter(is.na(no_press) | no_press == FALSE)
  }

  df %>%
    select(spelling, rt) %>%
    mutate(participant_id = file_id)
})


```


## Combine Data (Linguistic Data and Participants Data)

```{r}
combined_data <- all_data %>%
  left_join(word_info %>% select(spelling, Length, Zipf, is_word, is_word_binary, is_complex, morph), by = "spelling")

```


## Summary

```{r}
library(dplyr) #(Wickham et al., 2023)

# Summarize overall counts
summary_overall <- combined_data %>%
  summarise(
    n_participants    = n_distinct(participant_id),
    n_items           = n(),
    n_words           = sum(is_word_binary == 1),
    n_pseudowords     = sum(is_word_binary == 0)
  )


# Summarize by word vs. pseudoword
summary_by_type <- word_info %>%
  mutate(type = if_else(is_word_binary == 1, "word", "pseudoword")) %>%
  group_by(type) %>%
  summarise(
    n_items        = n(),
    n_complex      = sum(is_complex),
    n_simple       = sum(1L - is_complex),
    avg_complexity = mean(morph),
    min_complexity = min(morph),
    max_complexity = max(morph),
    avg_length     = mean(Length),
    min_length     = min(Length),
    max_length     = max(Length),
    avg_zipf       = mean(Zipf, na.rm = TRUE),
    min_zipf       = min(Zipf, na.rm = TRUE),
    max_zipf       = max(Zipf, na.rm = TRUE)
  )


summary_overall

print(summary_by_type, n = Inf, width = Inf)

```

# Linear Mixed Model

```{r}
#model accounts for individual variability, (1 | participant_id)
#Fixed effects: Zipf, Length, is_word_binary, is_complex 
model <- lmer(rt ~ Zipf + Length + is_word_binary + is_complex + (1 | participant_id), data = combined_data)

suppressWarnings(summary(model))

```

### Since there is high negative correlation between is_word_binary and Zipf (all pseudowords have frequency of 0 and therefore Zipf of 1.3555), other evaluations will be made on words and pseudowords separately

```{r}
# Subset of real words 
real_words <- combined_data %>% filter(is_word_binary == 1)
#Subset of pseudowords
pseudowords <- combined_data %>% filter(is_word_binary == 0)

# Real words model (including zipf, is_complex)
real_words_model <- lmer(rt ~ Zipf + Length + is_complex + (1 | participant_id), data = real_words)
suppressWarnings(summary(real_words_model))

# Real words model (including zipf)
real_words_model2 <- lmer(rt ~ Zipf + Length  + (1 | participant_id), data = real_words)
summary(real_words_model2)

# Pseudowords model (excluding Zipf, including is_complex)
pseudowords_model <- lmer(rt ~ Length + is_complex + (1 | participant_id), data = pseudowords)
summary(pseudowords_model)

# Pseudowords model (excluding Zipf)
pseudowords_model2 <- lmer(rt ~ Length + (1 | participant_id), data = pseudowords)
summary(pseudowords_model2)

```

# Plots

## Participants

```{r}
ranef_data <- ranef(model)$participant_id
ranef_data$participant_id <- rownames(ranef_data)

ggplot(ranef_data, aes(x = reorder(participant_id, `(Intercept)`), y = `(Intercept)`)) +
  geom_point(color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Random Intercepts per Participant",
       x = "Participant ID",
       y = "Random Intercept Estimate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```




```{r}
# QQ plot for residuals to check normality
residuals <- residuals(model)
qqnorm(residuals)
qqline(residuals, col = "red")
```

```{r}
# Plot random intercepts for participants
ranef_data <- as.data.frame(ranef(model)$participant)
ggplot(ranef_data, aes(x = (1:nrow(ranef_data)), y = `(Intercept)`)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Random Intercepts for Participants", x = "Participant", y = "Random Intercept (Effect on RT)") +
  theme_minimal()
```

## Length

```{r}

ggplot(combined_data, aes(x = Length, y = rt, color = as.factor(is_word_binary))) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("0" = "pink", "1" = "purple"),
                     labels = c("Pseudoword", "Real word"),
                     name = "Type") +
  labs(title = "Reaction Time vs Length",
       x = "Length of String",
       y = "Reaction Time (RT)") +
  theme_minimal()
```

```{r}
# RT vs Length by Word Type
ggplot(combined_data, aes(x = Length, y = rt)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", se = TRUE, color = "red") +
  facet_wrap(~ is_word) +
  labs(title = "Effect of Word Length on RT by Word Type",
       x = "Length", y = "Reaction Time (s)") +
  theme_minimal()



```

```{r}
ggplot(combined_data, aes(x = Length, y = rt, group = participant_id)) +
  geom_smooth(method = "lm", se = FALSE, aes(color = participant_id), show.legend = FALSE) +
  labs(title = "RT vs Length per Participant",
       x = "Length",
       y = "Reaction Time (RT)") +
  theme_minimal()

```

## Zipf

```{r}
ggplot(real_words, aes(x = Zipf, y = rt)) +
  geom_point(alpha = 0.2, color = "purple") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(title = "Reaction Time vs Zipf Frequency (Real Words)",
       x = "Zipf Frequency",
       y = "Reaction Time (s)") +
  theme_minimal()
```

## Pseudoword vs word

```{r}
ggplot(combined_data, aes(x = Length, y = rt*1000)) +
  geom_point(alpha = 0.2) +
  
 
  stat_summary(aes(group = 1),
             fun = mean, geom = "point",
             colour = "pink", size = 2)+
  stat_summary(aes(group = 1), fun = median,
             geom = "line", linetype = "dashed",
             colour = "blue", size = 0.9) +
stat_summary(aes(group = 1), fun = median,
             geom = "point", colour = "blue", size = 2)+

  
  ## red LOESS smooth
  geom_smooth(method = "loess", se = TRUE, colour = "red") +
  
  facet_wrap(~ is_word) +
  labs(title = "Effect of Word Length on RT by Word Type",
       x = "Number of letters", y = "Reaction time (ms)") +
  theme_minimal()


```

```{r}
ggplot(combined_data, aes(x = as.factor(is_word_binary), y = rt, fill = as.factor(is_word_binary))) +
  geom_boxplot() +
  scale_fill_manual(values = c("0" = "pink", "1" = "purple"),
                    labels = c("Pseudoword", "Real word"),
                    name = "Type") +
  scale_x_discrete(labels = c("Pseudoword", "Real word")) +
  labs(title = "Reaction Times by Word Type",
       x = "Type",
       y = "Reaction Time (RT)") +
  theme_minimal()
```

## is_complex

```{r}
ggplot(combined_data,
       aes(x = as.factor(is_complex),
           y = rt,
           fill = as.factor(is_complex))) +
  geom_boxplot(alpha = .8) +
  scale_fill_manual(values = c("0" = "skyblue", "1" = "steelblue"),
                    labels  = c("Simple", "Complex"),
                    name    = "Complexity") +
  scale_x_discrete(labels = c("Simple", "Complex")) +
  labs(title = "Reaction Times by Morphological Complexity",
       x = "Complexity",
       y = "Reaction Time (RT)") +
  theme_minimal()
```

```{r}
ggplot(combined_data,
       aes(x = morph,
           y = rt)) +
  geom_point(alpha = 0.25, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "black") +
  labs(title = "Reaction Time vs. Morpheme Count",
       x = "Number of Morphemes",
       y = "Reaction Time (RT)") +
  theme_minimal()

```

